---
title: "IE590_Project_Proposal"
author: "Lakshman , Utkuhan, Suraj, Kendrick" 
date: "3/26/2021"
output:
  html_document: default
  pdf_document: default
---

```{r saving Rdata}
save(list=ls(all=T),file='IE590_LR_ProjectProposal_0326.RData')
```

```{r message=FALSE}
#this code block is to load required libraries
#install.packages("readxl")
library(readxl)
library(ggplot2)
library(dplyr)
library(gridExtra)
library(corrplot)
library(car)
library(glmnet)
library(MASS)
library(ISLR)
library(gam)
library(splines)
```

## Exploratory Data Analysis
```{r importing data}
evdata = read_excel("EVDataFile2011-2019.xlsx")
evdata = as.data.frame(evdata)
attach(evdata)
```
The data contains 450 observations of 34 variables

```{r basic view}
colnames(evdata)[28] = "LAND_AREA"
head(evdata)
str(evdata)
summary(evdata)

```
Findings:
All of the columns, except for State, appear to be numeric or qualitative.
POP, LandArea,POPDENS are all interrelated so there would be a lot of correlation
DETOUSE, DETHOUSE%, OOH, OOH% are also very correlated
VEHMFG state manufacturing in millions of dollars

```{r checking if there are any NAs or missing values in the data}
#checking for NAs 

for (i in 1: ncol(evdata)){
  if (sum(complete.cases(evdata[,i])) != nrow(evdata)){
    print(paste("The column number" , i, "has NA values."))
    print(paste("The name of the column is", colnames(evdata)[i]  ))
    print(paste("Number of NA values: ", nrow(evdata) - sum(complete.cases(evdata[,i])) ))
    #evdata = evdata[-which(complete.cases(evdata[,i]) == FALSE),]
  }
  
}

#removing the values that have NA in them for WIND
evdata = evdata[which(complete.cases(evdata$WIND) == TRUE),]

#removing the values that have NA in them for CO2 and GAS
evdata = evdata[which(complete.cases(evdata$GAS) == TRUE),]
evdata = evdata[which(complete.cases(evdata$CO2) == TRUE),]

```
We already know the information for columns CO2 emissions and annual gas prices are not available for
the month of 2019. This corresponds to 50 of the rows or values.

What are the different kinds of plots that can be drawn?
Response vs all of the predictors
Spread of all of the predictors
Pair wise
Correlation check

sales per year
sales per state
sales per number of stations per year
sales vs other predictors / wrapped per year

```{r response plot}
x = evdata$SALES
x[which(x == 0)] = 1E-9

par(mfrow = c(1,2))
hist(x, breaks = 100, main = 'Histogram of Sales', xlab = 'Sales')
h = hist(log(x), breaks = 100, main = 'Histogram of log Sales', xlab = 'log Sales')
xfit <- seq(min(log(x)), max(log(x)), length = 1000)
yfit = pnorm(xfit, mean = mean(x), sd = sd(x))
yfit = yfit*diff(h$mids[1:2])*length(x)
lines(xfit, yfit)

par(mfrow = c(1,2))
plot(density(x), main = 'Density plot of Sales')
plot(density(log(x)), main = 'Density plot of log Sales')
```

Converted "0" values for Sales to 1e-9 in case log transformation is needed for the model. 

```{r checking for individual graps}
#pairs(SALES~.- State, data = evdata)

sales_year = evdata %>% group_by(YEAR) %>% summarise(sales = sum(SALES))

ggplot(data = sales_year, mapping = aes(x = YEAR, y = sales))+
  geom_line()+
  geom_text(aes(label = sales), nudge_y = 7000) +
  labs(x = 'Year', y = 'Sales of EVs', title = 'Sales of EVs vs Year')

sales_state = evdata %>% group_by(STATE) %>% summarise(sales = sum(SALES))
ggplot(data= sales_state, mapping = aes(x = reorder(STATE, -sales), y = sales)) +
  geom_col()+
  labs(x = 'State', y = 'Sales of EVs', title = 'Sales of EVs by State')+
  theme(axis.text.x = element_text(angle = 90))

  

```


Findings:
as can be seen in the statewise sales data CA dwarfs every other state
in its EV sales over the years from 2011 to 2018


This section is to generate plots for response (SALES) with each of the predictors
excluding State
```{r plotting response with each of the predictors}
#writing loops to plot sales vs each of the predictor variables
#32 plots except for State and SALES. Splitting this as plots of 4


# ggplot(data = evdata, mapping = aes_string(y = 'SALES', x = colnames(evdata)[i]))+
#       geom_point() +
#       labs(x = colnames(evdata)[i], y = 'Sales of EVs', title = paste('Sales of EVs vs', colnames(evdata)[i])  )
# 
ev_plot = evdata[,c(-1,-3)]

for(j in seq(1,32, 4)){
  par(mfrow = c(2,2))
  for (i in seq(j, min(j+3, ncol(ev_plot))  )  ){
    if(colnames(ev_plot)[i] == 'SALES') {
      next
      }
    else{
      plot(ev_plot[, colnames(ev_plot)[i]], ev_plot$SALES, 
           main = paste('EV Sales vs', colnames(ev_plot)[i]), xlab =
             colnames(ev_plot)[i], ylab = 'Sales')
      }
    }
  }
  
```



```{r plotting pairs, echo = FALSE}

#plotting pairs as too many variables to be of any use


#ev_pairs = evdata[, which(colnames(evdata) != 'State')]
#par(mfrow = c(1,1))
#par(mar = c(0.1,0.1,0.1,0.1))
#pairs(SALES~., data = ev_pairs[,1:10])



```


```{r checking for cor}

ev_wo_sales = ev_plot[, which(colnames(ev_plot) != 'SALES')]
ev_cor = cor(ev_wo_sales) 
#ev_cor[lower.tri(ev_cor, diag = TRUE)] = NA
corrplot(ev_cor, method = 'circle', type = 'upper',
         tl.srt = 65, tl.cex = 0.7, 
         title = 'Cor Plot Showing All Predictor Correlations', 
         mar = c(0,0,2,0))

#writing a function to perform hypothesis testing on correlation
#using the cor.test function
cor_pval = cor.mtest(ev_cor)$p
colnames(cor_pval)<- rownames(cor_pval)<- colnames(ev_wo_sales)

#this displays the colors of significant correlation
corrplot(ev_cor, type = 'upper',
         tl.srt = 65, tl.cex = 0.7, p.mat = cor_pval, sig.level = 0.01, diag = FALSE,
         insig = 'blank', 
         title = 'Cor Plot Showing Sig Correlations With alpha = 0.01', 
         mar = c(0,0,2,0))


#ev_cor_high = sapply(ev_cor, function(x) ifelse (abs(x) >= 0.75, x, "NA"))
ev_cor_high = ev_cor
#ev_cor_high = ev_cor_high[ev_cor_high > 0.75]
ev_cor_high[abs(ev_cor_high) < 0.75] = NA
ev_cor_high = as.matrix(ev_cor_high)

corrplot(ev_cor_high, type = 'upper',
         tl.srt = 65, tl.cex = 0.7, diag = FALSE, na.label = "  ", 
         title = 'Cor Plot Showing Correlations >= 0.75', 
         mar = c(0,0,2,0))


#displaying correlations of those with significant correlation
#indices gives data frame or pair of predictors that have >0.75 correlation
indices = which(cor_pval < 0.05, arr.ind = TRUE)
indices = as.data.frame(indices)
indices$cor = numeric(length = dim(indices)[1])

for (i in 1:dim(indices)[1]){
  indices$cor[i] = ev_cor[as.numeric(indices[i,'row']), as.numeric(indices[i,'col'])]
}
#indices$cor = ev_cor[as.numeric(indices[,'row']), as.numeric(indices[,'col'])]

indices[, 'row'] = rownames(cor_pval)[as.numeric(indices[, 'row'])]
indices[, 'col'] = colnames(cor_pval)[as.numeric(indices[, 'col'])]

len = nrow(indices)
indices$rep = 'first'

for (i in seq(1:nrow(indices))){
  if (indices[i,'row'] == indices[i, 'col']){
    indices$rep[i] = 'duplicate'
  }
}

for (i in seq(1,(len - 1)) ){
  for (j in seq((i+1),len)){
    if( (indices[i, 'row'] == indices[j, 'col']) & (indices[i, 'col'] == indices[j, 'row'])){
      indices$rep[i] = 'duplicate'
    }
  }
}

#cleaning it up to remove extra columns, SALES, and cor with values less than 0.75
indices = indices[which(indices$rep != 'duplicate'),]
indices = indices[which(indices$cor >= 0.75), ]
indices = indices[which(indices$row != 'SALES'), ]
indices = indices[, which(colnames(indices) != 'rep')]
indices
```
indices gives list of all pairs where the correlation is high
corr plot shows the significant correlations, taken for alpha = 0.01

## Linear Models
```{r full linear model}
set.seed(1)
evdata$SALES[which(evdata$SALES == 0)] = 1E-9
sample_rows <- sample(length(evdata[,1]),size = .8*length(evdata[,1]),replace = F)



evdata.train <- evdata[sample_rows,]
evdata.test <- evdata[-sample_rows,]

#Model 1
full_lin_model <- lm(SALES~.-STATE-YEAR, data = evdata.train)
summary(full_lin_model)


full_lin_model.test <- predict(full_lin_model,evdata.test)
print(paste("Full Linear Model MSE:" , round(mean((evdata.test$SALES - full_lin_model.test)^2),2)))

vif(full_lin_model)

#Stepwise Function
red_lin_model <- step(full_lin_model, direction = "both", trace = 0)
summary(red_lin_model)

red_lin_model.test <- predict(red_lin_model,evdata.test)
print(paste("Reduced Linear Model MSE:", round(mean((evdata.test$SALES - red_lin_model.test)^2),2)))


res <- rstandard(red_lin_model)
par(mfrow = c(2,3))

#histogram of residuals
hist(res, prob = 'T', col = 'gray', main = 'Histogram of Residuals', 
     xlab = 'Residuals')
lines(density(res), col = 'red', lwd = 3)

#norm prob plot of residuals 
qqnorm(res,ylab = 'Residuals', main = 'Normal Probability Plot', pch = 19, 
       col = 'blue', xlab = 'Theoretical Quantiles')
qqline(res)

#residuals vs order plot
plot(res , xlab = 'Observation Order', ylab = 'Residuals', pch = 19, 
     col  ='blue', main = 'Residuals vs Order')
abline(0,0,lwd=3, col = 'red')

#Residuals vs Fitted plot
plot(res~predict(red_lin_model), pch = 19, col = 'blue', main = 'Residuals vs Fitted', 
     ylab = 'Residuals', xlab = 'Fitted Values')
abline(0,0, lwd= 3 , col = 'red')

#Observed vs fitted plot
plot(predict(red_lin_model),evdata.train$SALES, pch = 19, col = 'blue', 
     main = 'Observed vs Fitted', ylab = 'Observed Values', 
     xlab = 'Fitted Values')
abline(0,1,lwd=3,col ='red')



bc <- boxcox(red_lin_model)

```

[1] "Full Linear Model MSE: 1894500.64"

[1] "Reduced Linear Model MSE: 1918229.66"

```{r transformed linear model}
#Transformed Linear Model 
lambda <- bc$x[which.max(bc$y)]

trans_full_lin_model <- lm(((SALES^lambda-1)/lambda)~.-STATE-YEAR, data = evdata.train)
summary(trans_full_lin_model)
trans_full_lin_model.test <- predict(trans_full_lin_model,evdata.test)

print(paste("Full Transformed Linear Model MSE:" , round(mean((evdata.test$SALES - trans_full_lin_model.test)^2),2)))

#Reduced Transformed Linear Model
trans_red_lin_model <- step(trans_full_lin_model, direction = "both", trace = 0)
summary(trans_red_lin_model)
trans_red_lin_model.test <- predict(trans_red_lin_model,evdata.test)

print(paste("Reduced Transformed Linear Model MSE:", round(mean((evdata.test$SALES - trans_red_lin_model.test)^2),2)))

res_trans <- rstandard(trans_red_lin_model)
par(mfrow = c(2,3))
#histogram of residuals
hist(res_trans, prob = 'T', col = 'gray', main = 'Histogram of Residuals', 
     xlab = 'Residuals')
lines(density(res_trans), col = 'red', lwd = 3)

#norm prob plot of residuals 
qqnorm(res_trans,ylab = 'Residuals', main = 'Normal Probability Plot', pch = 19, 
       col = 'blue', xlab = 'Theoretical Quantiles')
qqline(res)

#residuals vs order plot
plot(res_trans , xlab = 'Observation Order', ylab = 'Residuals', pch = 19, 
     col  ='blue', main = 'Residuals vs Order')
abline(0,0,lwd=3, col = 'red')

#Residuals vs Fitted plot
plot(res_trans~predict(trans_red_lin_model), pch = 19, col = 'blue', main = 'Residuals vs Fitted', 
     ylab = 'Residuals', xlab = 'Fitted Values')
abline(0,0, lwd= 3 , col = 'red')

#Observed vs fitted plot
plot(predict(trans_red_lin_model),evdata.train$SALES, pch = 19, col = 'blue', 
     main = 'Observed vs Fitted', ylab = 'Observed Values', 
     xlab = 'Fitted Values')
abline(0,1,lwd=3,col ='red')

```

We used boxcox method to transform our response variable. It seems to fix the normality problem. 

[1] "Full Transformed Linear Model MSE: 14495637.34"

[1] "Reduced Transformed Linear Model MSE: 14495640.7"

## Ridge and Lasso Regression
```{r ridge and lasso }
x <- as.matrix(evdata.train[,-1:-3])
y <- as.matrix(evdata.train$SALES)

#Ridge Regression
ridge_model <- glmnet(x,y = y,family = "gaussian", alpha = 0)
ridge_model.test <- predict(ridge_model,x,s = 1)
print(paste("Ridge Regression MSE:" ,round(mean((evdata.train$SALES - ridge_model.test)^2),2)))

cv.out.r<-cv.glmnet(x,y,alpha=0)
plot(cv.out.r)

ridge_model_cv.test <- predict(ridge_model,x,s = cv.out.r$lambda.mi)
print(paste("Ridge Regression(CV) MSE:" ,round(mean((evdata.train$SALES - ridge_model_cv.test)^2),2)))

#Lasso Regression
lasso_model <- glmnet(x,y = y,family = "gaussian", alpha = 1)
lasso_model.test <- predict(lasso_model,x,s = 1)
print(paste("Lasso Regression MSE:" ,round(mean((evdata.train$SALES - lasso_model.test)^2),2)))

cv.out.l<-cv.glmnet(x,y,alpha=1)
plot(cv.out.l)

lasso_model_cv.test <- predict(ridge_model,x,s = cv.out.l$lambda.mi)
print(paste("Lasso Regression(CV) MSE:" ,round(mean((evdata.train$SALES - lasso_model_cv.test)^2),2)))
```

## Generalized Additive Models
```{r Fitting a GAM model excluding State & Year}
gam.1 <- gam(SALES~-STATE-YEAR,data = evdata.train)
gam.1.test <- predict(gam.1,evdata.test)
print(paste("simple GAM MSE:", round(mean((evdata.test$SALES - gam.1.test)^2),2)))
```


```{r GAM using all variables}
gam.full <- gam(SALES~-STATE-YEAR,  data = evdata.train)
gam.full.test <- predict(gam.full,evdata.test)
print(paste("full GAM MSE:", round(mean((evdata.test$SALES - gam.full.test)^2),2)))

scope_list = list(
  "SOLAR" = ~1 + SOLAR + s(SOLAR, df=2) + s(SOLAR, df=3) + s(SOLAR, df =4) +        s(SOLAR, df=5),
  "EDU"= ~1+ EDU+ s(EDU, df=2)+s(EDU, df=3)+s(EDU, df =4)+s(EDU, df=5),
"MEDINC"= ~1+ MEDINC+ s(MEDINC, df=2)+s(MEDINC, df=3)+s(MEDINC, df =4)+s(MEDINC, df=5),
"TIME"= ~1+ TIME+ s(TIME, df=2)+s(TIME, df=3)+s(TIME, df =4)+s(TIME, df=5),
"EMPLOY"= ~1+ EMPLOY+ s(EMPLOY, df=2)+s(EMPLOY, df=3)+s(EMPLOY, df =4)+s(EMPLOY, df=5),
"PRIVEH"= ~1+ PRIVEH+ s(PRIVEH, df=2)+s(PRIVEH, df=3)+s(PRIVEH, df =4)+s(PRIVEH, df=5),
"ELEC"= ~1+ ELEC+ s(ELEC, df=2)+s(ELEC, df=3)+s(ELEC, df =4)+s(ELEC, df=5),
"POP"= ~1+ POP+ s(POP, df=2)+s(POP, df=3)+s(POP, df =4)+s(POP, df=5),
"POVERTY"= ~1+ POVERTY+ s(POVERTY, df=2)+s(POVERTY, df=3)+s(POVERTY, df =4)+s(POVERTY, df=5),
"EVMOD"= ~1+ EVMOD+ s(EVMOD, df=2)+s(EVMOD, df=3)+s(EVMOD, df =4)+s(EVMOD, df=5),
"RANGE"= ~1+ RANGE+ s(RANGE, df=2)+s(RANGE, df=3)+s(RANGE, df =4)+s(RANGE, df=5),
"ZEM"= ~1+ ZEM+ s(ZEM, df=2)+s(ZEM, df=3)+s(ZEM, df =4)+s(ZEM, df=5),
"POPDENS"= ~1+ POPDENS+ s(POPDENS, df=2)+s(POPDENS, df=3)+s(POPDENS, df =4)+s(POPDENS, df=5),
"`DETACH%`"= ~1+ `DETACH%`+ s(`DETACH%`, df=2)+s(`DETACH%`, df=3)+s(`DETACH%`, df =4)+s(`DETACH%`, df=5),
"VMT"= ~1+ VMT+ s(VMT, df=2)+s(VMT, df=3)+s(VMT, df =4)+s(VMT, df=5),
"VEHMFG"= ~1+ VEHMFG+ s(VEHMFG, df=2)+s(VEHMFG, df=3)+s(VEHMFG, df =4)+s(VEHMFG, df=5),
  "WIND" = ~1 + WIND + s(WIND, df=2) + s(WIND, df=3) + s(WIND, df =4) + s(WIND,    df=5),
  "SGDP" = ~1 + SGDP + s(SGDP, df=2) + s(SGDP, df=3) + s(SGDP, df =4) + s(SGDP,    df=5),
  "CO2" = ~1 + CO2 + s(CO2, df=2) + s(CO2, df=3) + s(CO2, df =4) + s(CO2,          df=5),
  "PUBTR" = ~1 + PUBTR + s(PUBTR, df=2) + s(PUBTR, df=3) + s(PUBTR, df =4) +       s(PUBTR,df=5),
"TOTVEH" = ~1 + TOTVEH + s(TOTVEH, df=2) + s(TOTVEH, df=3) + s(TOTVEH, df =4) + s(TOTVEH,df=5),
"GAS" = ~1 + GAS + s(GAS, df=2) + s(GAS, df=3) + s(GAS, df =4) + s(GAS,df=5),
"PORTS" = ~1 + PORTS + s(PORTS, df=2) + s(PORTS, df=3) + s(PORTS, df =4) + s(PORTS,df=5),
  "STATIONS" = ~1 + STATIONS + s(STATIONS, df=2) + s(STATIONS, df=3) +             s(STATIONS, df =4) + s(STATIONS,df=5),
  "INCENT" = ~1 + INCENT + s(INCENT, df=2) + s(INCENT, df=3) +             s(INCENT, df =4) + s(INCENT,df=5),
  "OOH" = ~1 + OOH + s(OOH, df=2) + s(OOH, df=3) + s(OOH, df =4) + s(OOH,df=5),
  "DETACH" = ~1 + DETACH + s(DETACH, df=2) + s(DETACH, df=3) +             s(DETACH, df =4) + s(DETACH,df=5),
  "CHARGE" = ~1 + CHARGE + s(CHARGE, df=2) + s(CHARGE, df=3) +             s(CHARGE, df =4) + s(CHARGE,df=5), 
"LAND_AREA" = ~1 + LAND_AREA + s(LAND_AREA, df=2) + s(LAND_AREA, df=3) +             s(LAND_AREA, df =4) + s(LAND_AREA,df=5),
"`OOH%`" = ~1 + `OOH%` + s(`OOH%`, df=2) + s(`OOH%`, df=3) +             s(`OOH%`, df =4) + s(`OOH%`,df=5),
"PRICE" = ~1 + PRICE + s(PRICE, df=2) + s(PRICE, df=3) + s(PRICE, df =4) + s(PRICE,df=5)
)

gam_red <- step.Gam(gam.full,scope_list)

gam_final <- gam(SALES ~ s(SOLAR, df = 4) + EDU + s(TIME, df = 5) + s(ELEC, df = 5) +      EVMOD + s(`DETACH%`, df = 4) + SGDP + s(CO2, df = 2) + s(PORTS,      df = 5) + STATIONS + INCENT + s(`OOH%`, df = 5) + PRICE, data = evdata.train)
anova(gam.full,gam_final)

par(mfrow=c(1,10))
plot(gam_final, se=T)

gam_model.test <- predict(gam_final,evdata.test)

print(paste("full GAM MSE:", round(mean((evdata.test$SALES - gam.full.test)^2),2)))
print(paste("Reduced GAM MSE:", round(mean((evdata.test$SALES - gam_model.test)^2),2)))

```

